Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-06-12 pkg/paper/Paper/draft.Rnw
update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@118 b679b2f7-6539-45d0-a605-5ddcb976cde9

\renewcommand{\abstractname}{Summary}
\Abstract{
  To assess the effects of the EMU on inflation rate dynamics of
  its member states, the inflation rate series for 21~European countries
  are investigated for structural changes. To capture changes in mean,
  variance, and skewness of inflation rates, a generalized logistic
  model is adopted and complemented with structural break tests and
  breakpoint estimation techniques. These reveal considerable differences
  in the patterns of inflation dynamics and the structural changes
  therein. Overall, there is a convergence towards a lower mean inflation rate
  with reduced skewness, but accompanied by an increase in variance.
}

@@

As the way towards the EMU essentially consisted of three stages (I:~1990--1994,
a phase of liberalization; II: 1995--1998, a phase of convergence; III: 1999--2001,
a transition period ending with the introduction of the euro as legal tender),
it seems natural to ask whether this transition was accompanied by structural changes
in the inflation dynamics in the different European countries. Furthermore, a
possible fourth stage is the continuing integration of new member states, mostly
former communist countries. Typically, changes in the mean level of the
inflation rates are of most interest but there are also reasons for investigating
changes in higher moments, such as variance and skewness. \citet{emerson} emphasize that
a high inflation rate is usually also more variable and uncertain and thus
causes more relative price variability, leading to a less stability and a less
efficient price mechanism. Furthermore, \cite{jaro} finds a positive correlation
between the level and the standard deviation of inflation.

@@

Therefore, a thorough investigation of the evolution of the inflation dynamics
within a country over time requires assessment not only of the mean level but
also higher moments. Hence, we contribute to this field of research by
developing a new method for testing for structural breaks based on a generalized
logistic (GL) distribution. This provides flexible modeling of means, variances,
and skewness, while also allowing for fat tails. Structural breaks are tested
using score-based tests \citep{Andrews:1993,Zeileis:2005} and breakpoints are estimated
by extending the least squares approach of \cite{bp03} to (quasi-)maximum likelihood
estimation of GL distributions \citep[similar in spirit to the Gaussian model
employed by][]{Zeileis+Shah+Patnaik:2010}.

The empirical results show that there are different patterns and timings of
structural breaks in the inflation dynamics of different groups of countries.
Inflation rates in the core euro area did not change significantly after
completion of the first convergence stage, yielding low mean and variance.
The Southern euro area member states (and Ireland), on the other hand, managed
to decrease the mean level of inflation in the mid-1990s, but at the price
of higher volatility. Finally, the Eastern European countries reduced both mean
and variance on their way towards joining the EMU. Generally, the inflation rates
converged from very different distributions at the beginning of the sample to
rather similar patterns at the end of the 2000s. Skewness also plays an
important role in this convergence: Some Southern and Eastern countries
have clear breaks from right-skewed to roughly symmetric inflation rates,
signalling a reduction of size and frequency of outlying high inlation rates.

@@

The countries in this sample can be divided into three different groups: (1)~euro
countries -- Austria, Belgium, Estonia, France, Finland, Germany, Greece,
Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia,
Spain; (2)~EU members not participating in the ERM~II -- Czech Republic,
Hungary, Poland, United Kingdom, Sweden; (3)~Denmark which stands on its own as a
member of the EU and the ERM~II, but not yet of the EMU. 
Latvia, Lithuania, Bulgaria, Romania are excluded due to data scarcity.
Cyprus, and Malta are not included since they are very small economies.

Seasonal adjustment is necessary due to the harmonization of the treatment
of sales prices\footnote{Inclusion of sales prices in the HICP was demanded by
a Commission regulation \citep[see][for details]{eurlex}.} in the HICP, resulting in
changes of the seasonal inflation patters over time within several countries
and also between the countries. See \citet{pltm} for a detailed discussion.

@@

The goal of the modeling strategy, that is subsequently introduced, is the
following: detection of structural changes in the distributional properties of
the inflation rates over time, e.g., as potentially caused by interventions.
However, unlike many standard least squares approaches \citep[such as][]{bp03},
not only changes in the mean level should be addressed but also changes in
variance and potential skewness.

@@

In the following, details are provided about (1)~the GL distribution and its properties,
(2)~how it can be estimated (under parameter stability), (3)~how parameter stability
can be assesed, and (4)~how breakpoints can be estimated in the presence of parameter
instability.

@@

The GL distribution employed here corresponds to the type~1 GL distribution 
from\citet{johnson}, using the parametrization and inference as in \citet{Shao:2002}. It 
has the log-density function
%
\begin{eqnarray}
%% f(y ~|~ \theta, \sigma, \delta) & = &
%% \frac{\frac{\delta}{\sigma} \cdot \exp\{-\frac{y - \theta}{\sigma}\}}{(1+\exp\{-\frac{y - \theta}{\sigma}\})^{(\delta+1)}} 
\ell(y ~|~ \theta, \sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{y - \theta}{\sigma} -
                                         (\delta + 1) \cdot \log\left(1 + \exp\left\{-\frac{y - \theta}{\sigma}\right\}\right)
\label{eq:dglogis}
\end{eqnarray}
%
with location~$\theta$, scale~$\sigma > 0$, and shape~$\delta > 0$. For $\delta = 1$ the
distribution simplifies to the logistic distribution, for $\delta < 1$ or $> 1$ it is
left- or right-skewed, respectively. The corresponding first three moments are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma(\cdot)$, $\gamma'(\cdot)$, and $\gamma''(\cdot)$ are the digamma function and its
first and second derivative, respectively.

@@

As usual in structural change analysis, a functional central limit theorem holds for
$\mathit{efp}(\cdot)$, which converges to a 3-dimensional Brownian
bridge: $\mathit{efp}(\cdot) ~\overset{d}{\rightarrow}~ W^0(\cdot)$. Based on
this various types of test statistics can be computed
(see \citealp{Zeileis:2005} for details). In the following, the 
sup$\mathit{LM}$ test of \cite{Andrews:1993} is employed, which performs particularly
well for single shift alternatives and if several of the three distribution
parameters change simultaneously. The test statistic with 10\% trimming is given by
%
\begin{eqnarray}
\sup_{t \in [0.1, 0.9]} \frac{\Vert \mathit{efp}(t) \Vert_2^2}{t(1-t)} \nonumber
\end{eqnarray}
%
and the appropriate $p$~values can be computed using from the corresponding
limiting distribution -- with $\mathit{efp}(t)$ replaced by $W^0(t)$.

@@

## Exchange Rate Mechanism information
## ERM~II, ERM~I (part of european monetary system expired in 1999-01-01 gave way to ERM II
## if never entered ""right?
## http://web.archive.org/web/20080101022356/http://lexikon.meyers.de/meyers/Europ%C3%A4isches_W%C3%A4hrungssystem
erm <- yearmon(c(
  Austria = 1995,
  Belgium = 1979 + 2/12,
  CzechRepubic = NA, 
  Denmark = 1979 + 2/12,
  Estonia = NA,
  Finland = 1996 + 9/12,
  France = 1979 + 2/12,
  Germany = 1979 + 2/12,
  Greece = 1998 + 2/12,
  Hungary = NA,
  Ireland = 1979 + 2/12,
  Italy = 1979 + 2/12,
  Luxembourg = 1979 + 2/12,
  Netherlands = 1979 + 2/12,
  Poland = NA,
  Portugal = 1992 + 3/12,
  Slovakia = NA,
  Slovenia = NA,
  Spain = 1986 + 5/12,
  Sweden = NA ,
  UK = NA,
  EU = NA
))

@@

To illustrate the proposed strategy, the case of Slovenia is considered. After a period of
relative instability following the independence from Yugoslavia, Slovenia was
successful in realigning its economy and introduced a number of reforms leading
to a stable growth in recent years. The good economic performance made it
possible for Slovenia to enter the ERM~II in June~2004 and later on to introduce
the euro in January~2007. The seasonally adjusted inflation series for Slovenia encompasses
the time from \Sexpr{start(x$data)} to \Sexpr{end(x$data)} and is depicted in the left
panel of Figure~\ref{fig:slovenia1}.

\begin{figure}[t!]
<<slovenia-plots1, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
plot(x$data, main = "", ylab = "Monthly inflation rate (in percent)", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), col = 4)
if(x$nbreaks > 0) lines(confint(x$breakpoints, breaks = x$nbreaks))
plot(x$efp, functional = supLM(0.1), main = "", ylab = "LM statistics")
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{\label{fig:slovenia1} Inflation rate series for Slovenia
with breakpoint estimate (\Sexpr{breakdates(x$breakpoints)}), associated confidence interval, and fitted mean
from generalized logistic distribution (left). Corresponding sup$\mathit{LM}$ test (right)
with sequence of $\mathit{LM}$~statistics,
critical value at 5\% level (horizontal red line line), and estimated breakpoint (vertical dashed line).}

<<slovenia-plots2, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
for(i in 1:2) plot(x$model[[i]], moments = TRUE,
  xlim = c(-1.5, 2.2), ylim = c(0, 1.3),
  breaks = seq(-1.5, 2, by = 0.25),
  xlab = paste(names(x$model)[i], "\n", "Goodness-of-fit p-value: ",
    format.pval(summary(x$model[[i]])$chisq.test$p.value, digits = 4),
    sep = ""))
@
\caption{\label{fig:slovenia2} Histogramm of observed monthly inflation rates in
Slovenia for \Sexpr{names(x$model)[1]} (left) and \Sexpr{names(x$model)[1]} (right),
along with fitted generalized logistic probability density function, associated
moments, and $\chi^2$ goodness-of-fit test.}
\end{figure}

Assuming a single stable set of parmaters (\ref{eq:h0}) is  not valid, as the
sequence of $\mathit{LM}$~statistics (see right panel of Figure~\ref{fig:slovenia1}) clearly
exceeds its 5\% critical value (red horizontal line), leading to a highly significant
$p$-value \Sexpr{spval[1]}. If a quadratic spectral kernal HAC estimator \citep{Andrews:1991}
would have been used to adjust (\ref{eq:efp}), the resulting $p$-value would be somewhat
larger with \Sexpr{spval[2]} but still clearly significant.

As there is evidence for at least one break, the LWZ criterion is employed for $B = 1, 2, \dots$
breakpoints with a minimal segment size of two years per segment. It assumes its minimum for $B = 1$
associated with the breakpoint \Sexpr{breakdates(x$breakpoints, breaks = x$nbreaks)}, depcited
by a vertical dashed line in both panels of Figure~\ref{fig:slovenia1}. The resulting segmented
fitted mean (horizontal blue lines in the left panel Figure~\ref{fig:slovenia1}) shows that
in before the break Slovenia experienced very high inflation rates but was successful in reducing
inflation to a much lower level afterwards.

@@

The estimated breakpoint can be matched very well with the timing of financial reforms
in Slovenia that severely decreased the annual growth in the M3 monetary aggregate.
These success of these actions ensured the Slovenian participation in
the ERM~II only a year later and the introduction of the euro two years later.

@@

As a first step towards the investigation of inflation convergence, the
structural breaks in the inflation dynamics within each countries are
investigated using the methods introduced in Section~\ref{sec:model}.
Table~\ref{tab:results1} reports the estimated breakpoints along with the
segment-specific moments (mean, variance, skewness) implied by the corresponding
parameter estimates. Furthermore, entry dates for ERM/ERM~II and the euro are
included as additional information. For ease of interpretation, countries with
similar patterns of change are grouped together and ordered alphabetically
within each group. If there is only a single segment, the sup$\mathit{LM}$ test
was non-significant at the 5\% level. In significant cases, the number of
breakpoints was selected via the LWZ criterion and subsequently all parameters
(breakpoints and segment-specific GL parameters) were estimated by ML. No
HAC~correction was employed as there was only negligible autocorrelation in the
empirical scores after inclusion of the breakpoints (if any).
In Table~\ref{tab:results1} the countries are grouped according to either a
similar timing of the changes in inflation dynamics or according to economic similarities, 
as in the case of the former communist countries in Eastern Europe.

@@

@@

The next group consists of three of the PIIGS countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successful in reducing its inflation rate, whereas Luxembourg
experienced a strong increase in its formerly very low inflation rate, both
in terms of mean and variances, also associated with a shift in skewness towards the
left. Apart from Luxembourg, the countries that had a structural break during the second 
phase of convergence were also successful in reducing mean inflation rates.
After the changes in this group, all countries have rather similar mean inflation
rates, closer to the levels of the previous two groups.

@@

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
late 1990s or early 2000s. Czech Republic, Estonia and Hungary experienced a break in 1997/1998
resulting from efforts to curb in inflation by decreasing the growth of the money 
supply. The timing of the break thus indicates the timing of the biggest
transitionally shocks on the way towards free market economies. In Poland, this
transition took two more years. Slovakia and Slovenia are somewhat different: On one hand,
they started with lower mean inflation rates and on the other hand, their inflation dynamics
both changed roughly one year prior to their entry into the ERM~II.

@@

Given the results above for the individual countries, a natural question is
whether there is convergence on an aggregate level and, if so, which pattern it
follows. Hence, we aggregate the results from Table~\ref{tab:results1} for the
individual countries into three groups and at four time points to reflect the
general dynamics of European inflation rates during the time period of interest.
The groups have been selected based on prior expectations (i.e., not based on
the breakpoints estimation results) as the ``core'' of the euro zone
(\Sexpr{paste(colnames(hicps)[euro], collapse = ", ")}), the PIIGS countries,
and the rest. Similarly, the time points have been chosen as to fall ``between''
times where transitions in one or more groups of countries could be expected:
Jan~\Sexpr{yrs[1]} during stage~I on the way towards the EMU, Jan~\Sexpr{yrs[2]}
during stage~II, Jan~\Sexpr{yrs[3]} after transition into the EMU but before the
financial crisis, and Jan~\Sexpr{yrs[4]} after the crisis. The result of these
considerations is presented in Table~\ref{tab:results2} which extracts the three
moments for each country from Table~\ref{tab:results1} at the indicated time
points and aggregates them within the group. Lower quartile, median, and upper
quartile (25\%, 50\%, 75\% percentiles) are provided to assess both changes in
location and spread across countries.

For the core euro countries, there are only changes from stage~I to stage~II
because after \Sexpr{yrs[2]} no breaks have been detected. The changes in this
first phase did not affect mean inflation much, while variance increased
slightly (in both quartiles) and skewness somewhat decreased (from slightly
right- to slightly left-skewed). The heterogeneity (assessed by the
interquartile range) decreased for all three moments of the inflation rates.
This reflects that the monetary policies of the corresponding countries were
already closely connected prior to stage~I and now even moved somewhat closer
together.

For the PIIGS countries, changes occured from \Sexpr{yrs[1]} to
\Sexpr{yrs[2]} and then again during the crisis, before \Sexpr{yrs[4]}.
In the first phase, mean inflation was almost halved, accompanied by a
doubling of variance and a change from clearly right-skewed to approximately
symmetric. The latter reflects that the occurence of shocks with high inflation
rates have been successfully reduced. Moreover, the changes in the PIIGS
countries between \Sexpr{yrs[3]} and \Sexpr{yrs[4]} lead to a further
decrease in mean inflation and associated skewness which are both caused by
several extremely low inflation (or even deflation) rates during the time
following the crisis.

The other countries do not exhibit any significant breakpoints during the
mid-1990s, but many changes occur in the transition period from \Sexpr{yrs[2]}
to \Sexpr{yrs[3]}, encompassing both the completion of transformation from
planned to free market economies and the intended entry into the EMU.
During this time, the countries were successful in
decreasing all three moments of their inflation distributions while
simultaneously decreasing the heterogeneity (as captured by the corresponding
interquartile ranges) across countries. Before the transition all three
moments were extremely large compared to the other two groups, while afterwards
all moments are much closer to those of the other groups. In Jan~\Sexpr{yrs[3]},
mean inflation is in between the levels for the core euro and PIIGS countres,
variance is still slightly higher, and skewness is not too far from symmetry.

In context of the convergence issue, we find that up to \Sexpr{yrs[3]}
the inflation dynamics of all three groups of countries become much
more similar, both within and across groups. The interquartile ranges
of the means decline and become tighter for all groups and there appears to
be convergence towards some common mean inflation rate in the
countries of our sample. Variances also move more closely together, corresponding
to an increase for the core euro and PIIGS countries but a clear decrease
for the others. Right-skewness (and thus large outlying inflation rates)
has been reduced for all three groups of countries but to different extents.
Thus, while ending in similar inflation dynamics, the timing and patterns of
changes in this convergence process are rather different across
the three groups, depending on the different starting situations of the
countries.

@@

The results from the previous sections convey that the financial
crisis of 2008/2009 did not have a significant effect on the inflation dynamics
in most of the European countries (except Ireland and Spain).
This may be somewhat surprising and is not enitrely correct. In fact,
there is a rather clear effect but as it is very short-lived and not
persistent it is not picked up by the structural break techniques.

>>>>

\renewcommand{\abstractname}{Summary}
\Abstract{
  To assess the effects of the EMU on inflation rate dynamics of
  its member states, the inflation rate series for 21~European countries
  are investigated for structural changes. To capture changes in mean,
  variance, and skewness of inflation rates, a generalized logistic
  model is adopted and complemented with structural break tests and
  breakpoint estimation techniques. These reveal considerable differences
  in the patterns of inflation dynamics and the structural changes
  therein. Overall, there is a convergence towards a lower mean inflation rate
  with reduced skewness, but it is accompanied by an increase in variance.
}

@@

As the way towards the EMU essentially consisted of three stages (I:~1990--1994,
a phase of liberalization; II: 1995--1998, a phase of convergence; III: 1999--2001,
a transition period ending with the replacement of the national currencies by the euro),
it seems natural to ask whether this transition was accompanied by structural changes
in the inflation dynamics in the different European countries. Furthermore, a
possible fourth stage is the continuing integration of new member states, mostly
former communist countries. Typically, changes in the mean level of the
inflation rates attract the widest interest, but there are also reasons for investigating
changes in higher moments, such as variance and skewness. \citet{emerson} emphasize that
a high inflation rate is usually also more variable and uncertain and thus
causes more relative price variability, leading to a less
efficient price mechanism. Furthermore, \cite{jaro} finds a positive correlation
between the level and the standard deviation of inflation.

@@

Therefore, a thorough investigation of the evolution of the inflation dynamics
within a country over time requires assessment not only of the mean level, but
also of higher moments. Hence, we contribute to this field of research by
developing a new method for testing for structural breaks based on a generalized
logistic (GL) distribution. This provides flexible modeling of means, variances,
and skewness, while also allowing for fat tails. Structural breaks are tested
using score-based tests \citep{Andrews:1993,Zeileis:2005} and breakpoints are estimated
by extending the least squares approach of \cite{bp03} to (quasi-)maximum likelihood
estimation of GL distributions \citep[similar in spirit to the Gaussian model
employed by][]{Zeileis+Shah+Patnaik:2010}.

The empirical results show that there are different patterns and timings of
structural breaks in the inflation dynamics of different groups of countries.
Inflation rates in the core euro area did not change significantly after
completion of the first convergence stage, yielding low mean and variance.
The Southern euro area member states (and Ireland), on the other hand, managed
to decrease the mean level of inflation in the mid-1990s, but at the price
of higher volatility. Finally, the Eastern European countries reduced both mean
and variance on their way towards joining the EMU. Generally, the inflation rates
converged from very different distributions at the beginning of the sample to
rather similar patterns at the end of the 2000s. Skewness also plays an
important role in this convergence: Some Southern and Eastern countries
have clear breaks from right-skewed to roughly symmetric inflation rates,
signaling a reduction of size and frequency of outlying high inflation rates.

@@

The countries in this sample can be divided into three different groups: (1)~euro
countries -- Austria, Belgium, Estonia, France, Finland, Germany, Greece,
Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia,
Spain; (2)~EU members not participating in the ERM~II -- Czech Republic,
Hungary, Poland, United Kingdom, Sweden; (3)~Denmark which stands on its own as a
member of the EU and the ERM~II, but not yet of the EMU. 
Latvia, Lithuania, Bulgaria, Romania are excluded due to data scarcity.
Cyprus, and Malta are not included since they are very small economies. 
%% and are also not members of the OECD

Seasonal adjustment is necessary due to the harmonization of the treatment
of sales prices\footnote{Inclusion of sales prices in the HICP was demanded by
a Commission regulation \citep[see][for details]{eurlex}.} in the HICP, resulting in
changes of the seasonal inflation patterns over time within several countries
and also between the countries. See \citet{pltm} for a detailed discussion.

@@

The goal of the modeling strategy, that is subsequently introduced, is the
following: detection of structural changes in the distributional properties of
the inflation rates over time, e.g., as potentially caused by interventions.
However, unlike many standard least squares approaches \citep[such as][]{bp03},
not only changes in the mean level should be addressed, but also changes in
variance and potential skewness.

@@

In the following, details are provided about (1)~the GL distribution and its properties,
(2)~how it can be estimated (under parameter stability), (3)~how parameter stability
can be assessed, and (4)~how breakpoints can be estimated in the presence of parameter
instability.

@@

The GL distribution employed here corresponds to the type~1 GL distribution 
from \citet{johnson}, using the parametrization and inference as in \citet{Shao:2002}. It 
has the log-density function
%
\begin{eqnarray}
%% f(y ~|~ \theta, \sigma, \delta) & = &
%% \frac{\frac{\delta}{\sigma} \cdot \exp\{-\frac{y - \theta}{\sigma}\}}{(1+\exp\{-\frac{y - \theta}{\sigma}\})^{(\delta+1)}} 
\ell(y ~|~ \theta, \sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{y - \theta}{\sigma} -
                                         (\delta + 1) \cdot \log\left(1 + \exp\left\{-\frac{y - \theta}{\sigma}\right\}\right)
\label{eq:dglogis}
\end{eqnarray}
%
with location~$\theta$, scale~$\sigma > 0$, and shape~$\delta > 0$. For $\delta = 1$ the
distribution simplifies to the logistic distribution, for $\delta < 1$ or $> 1$ it is
left- or right-skewed, respectively. The corresponding first three moments are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma(\cdot)$, $\gamma'(\cdot)$, and $\gamma''(\cdot)$ are the digamma function and its
first and second derivative, respectively.

@@

As usual in structural change analysis, a functional central limit theorem holds for
$\mathit{efp}(\cdot)$, which converges to a 3-dimensional Brownian
bridge: $\mathit{efp}(\cdot) ~\overset{d}{\rightarrow}~ W^0(\cdot)$. Based on
this, various types of test statistics can be computed
(see \citealp{Zeileis:2005} for details). In the following, the 
sup$\mathit{LM}$ test of \cite{Andrews:1993} is employed, which performs particularly
well for single shift alternatives and if several of the three distribution
parameters change simultaneously. The test statistic with 10\% trimming is given by
%
\begin{eqnarray}
\sup_{t \in [0.1, 0.9]} \frac{\Vert \mathit{efp}(t) \Vert_2^2}{t(1-t)} \nonumber
\end{eqnarray}
%
and the appropriate $p$~values can be computed from the corresponding
limiting distribution -- with $\mathit{efp}(t)$ replaced by $W^0(t)$.

@@

erm <- yearmon(c(
  Austria = 1995,
  Belgium = 1979 + 2/12,
  CzechRepubic = NA, 
  Denmark = 1979 + 2/12,
  Estonia = NA,
  Finland = 1996 + 9/12,
  France = 1979 + 2/12,
  Germany = 1979 + 2/12,
  Greece = 1998 + 2/12,
  Hungary = NA,
  Ireland = 1979 + 2/12,
  Italy = 1979 + 2/12,
  Luxembourg = 1979 + 2/12,
  Netherlands = 1979 + 2/12,
  Poland = NA,
  Portugal = 1992 + 3/12,
  Slovakia = NA,
  Slovenia = NA,
  Spain = 1986 + 5/12,
  Sweden = NA ,
  UK = NA,
  EU = NA
))

@@

To illustrate the proposed strategy, the case of Slovenia is considered. After a period of
relative instability following the independence from Yugoslavia, Slovenia was
successful in realigning its economy and introduced a number of reforms leading
to stable growth in recent years. The good economic performance made it
possible for Slovenia to enter the ERM~II in June~2004 and later on to introduce
the euro in January~2007. The seasonally adjusted inflation series for Slovenia encompasses
the time from \Sexpr{start(x$data)} to \Sexpr{end(x$data)} and is depicted in the left
panel of Figure~\ref{fig:slovenia1}.

\begin{figure}[t!]
<<slovenia-plots1, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
plot(x$data, main = "", ylab = "Monthly inflation rate (in percent)", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), col = 4)
if(x$nbreaks > 0) lines(confint(x$breakpoints, breaks = x$nbreaks))
plot(x$efp, functional = supLM(0.1), main = "", ylab = "LM statistics")
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{\label{fig:slovenia1} Inflation rate series for Slovenia
with breakpoint estimate (\Sexpr{breakdates(x$breakpoints)}), associated confidence interval, and fitted mean
from generalized logistic distribution (left). Corresponding sup$\mathit{LM}$ test (right)
with sequence of $\mathit{LM}$~statistics,
critical value at 5\% level (horizontal red line), and estimated breakpoint (vertical dashed line).}

<<slovenia-plots2, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
for(i in 1:2) plot(x$model[[i]], moments = TRUE,
  xlim = c(-1.5, 2.2), ylim = c(0, 1.3),
  breaks = seq(-1.5, 2, by = 0.25),
  xlab = paste(names(x$model)[i], "\n", "Goodness-of-fit p-value: ",
    format.pval(summary(x$model[[i]])$chisq.test$p.value, digits = 4),
    sep = ""))
@
\caption{\label{fig:slovenia2} Histogram of observed monthly inflation rates in
Slovenia for \Sexpr{names(x$model)[1]} (left) and \Sexpr{names(x$model)[1]} (right),
along with fitted generalized logistic probability density function, associated
moments, and $\chi^2$ goodness-of-fit test.}
\end{figure}

Assuming a single stable set of parameters as in (\ref{eq:h0}) is  not valid, since the
sequence of $\mathit{LM}$~statistics (see right panel of Figure~\ref{fig:slovenia1}) clearly
exceeds its 5\% critical value (red horizontal line), leading to a highly significant
$p$-value \Sexpr{spval[1]}. If a quadratic spectral kernel HAC estimator \citep{Andrews:1991}
would have been used to adjust (\ref{eq:efp}), the resulting $p$-value would be somewhat
larger with \Sexpr{spval[2]} but still clearly significant.

As there is evidence for at least one break, the LWZ criterion is employed for $B = 1, 2, \dots$
breakpoints with a minimal segment size of two years per segment. It assumes its minimum for $B = 1$
associated with the breakpoint \Sexpr{breakdates(x$breakpoints, breaks = x$nbreaks)}, depicted
by a vertical dashed line in both panels of Figure~\ref{fig:slovenia1}. The resulting segmented
fitted mean (horizontal blue lines in the left panel Figure~\ref{fig:slovenia1}) shows that
before the break, Slovenia experienced very high inflation rates but was successful in reducing
inflation to a much lower level afterwards.

@@

The estimated breakpoint can be matched very well with the timing of financial reforms
in Slovenia that severely decreased the annual growth in the M3 monetary aggregate.
The success of these actions ensured the Slovenian participation in
the ERM~II only a year later and the introduction of the euro two years afterwards.

@@

As a first step towards the investigation of inflation convergence, the
structural breaks in the inflation dynamics within each country are
investigated using the methods introduced in Section~\ref{sec:model}.
Table~\ref{tab:results1} reports the estimated breakpoints along with the
segment-specific moments (mean, variance, skewness) implied by the corresponding
parameter estimates. Furthermore, entry dates for ERM/ERM~II and the euro are
included as additional information. For ease of interpretation, countries with
similar patterns of change are grouped together and ordered alphabetically
within each group. If there is only a single segment, the sup$\mathit{LM}$ test
was non-significant at the 5\% level. In significant cases, the number of
breakpoints was selected via the LWZ criterion and subsequently all parameters
(breakpoints and segment-specific GL parameters) were estimated by ML. No
HAC~correction was employed as there was only negligible autocorrelation in the
empirical scores after inclusion of the breakpoints (if any).
In Table~\ref{tab:results1} the countries are grouped according to either a
similar timing of the changes in inflation dynamics or according to economic similarities, 
as in the case of the former communist countries in Eastern Europe.

@@

%% source for DM zone: http://ec.europa.eu/economy_finance/publications/publication6730_de.pdf

@@

The next group consists of three of the PIIGS countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successful in reducing its inflation rate, whereas Luxembourg
experienced a strong increase in its formerly very low inflation rate, both
in terms of mean and variance, also associated with a shift in skewness towards the
left. Apart from Luxembourg, the countries that had a structural break during the second 
phase of convergence were also successful in reducing mean inflation rates.
After the changes in this group, all countries have rather similar mean inflation
rates, which are closer to the levels of the previous two groups.

@@

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
late 1990s or early 2000s. Czech Republic, Estonia and Hungary experienced a break in 1997/1998
resulting from efforts to curb in inflation by decreasing the growth of the money 
supply. The timing of the break thus indicates the date when the the biggest
transitionally shocks on the way towards free market economies have been overcome. In Poland, this
transition took two more years. Slovakia and Slovenia are somewhat different: On the one hand,
they started with lower mean inflation rates and on the other hand, their inflation dynamics
both changed roughly one year prior to their entry into the ERM~II.

@@

Given the results above for the individual countries, a natural question is
whether there is convergence on an aggregate level and, if so, which pattern it
follows. Hence, we aggregate the results from Table~\ref{tab:results1} for the
individual countries into three groups and at four time points to reflect the
general dynamics of European inflation rates during the time period of interest.
The groups have been selected based on prior expectations (i.e., not based on
the breakpoints estimation results) as the ``core'' of the euro zone, the PIIGS countries,
and the rest. Similarly, the time points have been chosen as to fall ``between''
times where transitions in one or more groups of countries could be expected:
Jan~\Sexpr{yrs[1]} during stage~I on the way towards the EMU, Jan~\Sexpr{yrs[2]}
during stage~II, Jan~\Sexpr{yrs[3]} after transition into the EMU but before the
financial crisis, and Jan~\Sexpr{yrs[4]} after the crisis. The result of these
considerations is presented in Table~\ref{tab:results2} which extracts the three
moments for each country from Table~\ref{tab:results1} at the indicated time
points and aggregates them within the group. Lower quartile, median, and upper
quartile (25\%, 50\%, 75\% percentiles) are provided to assess both changes in
location and spread across countries.

For the core euro countries, there are only changes from stage~I to stage~II
because after \Sexpr{yrs[2]} no breaks have been detected. The changes in this
first phase did not affect mean inflation much, while variance increased
slightly (in both quartiles) and skewness decreased somewhat (from slightly
right- to slightly left-skewed). The heterogeneity (assessed by the
interquartile range) declined for all three moments of the inflation rates.
This reflects that the monetary policies of the corresponding countries were
already closely connected prior to stage~I and now even moved somewhat closer
together.

For the PIIGS countries, changes occurred from \Sexpr{yrs[1]} to
\Sexpr{yrs[2]} and then again during the crisis, before \Sexpr{yrs[4]}.
In the first phase, mean inflation was almost halved, accompanied by a
doubling of variance and a change from clearly right-skewed to approximately
symmetric. The latter reflects that the occurrence of shocks with high inflation
rates have been successfully reduced. Moreover, the changes in the PIIGS
countries between \Sexpr{yrs[3]} and \Sexpr{yrs[4]} lead to a further
decrease in mean inflation and associated skewness which are both caused by
several extremely low inflation rates during the time
following the crisis.

The other countries do not exhibit any significant breakpoints during the
mid-1990s, but many changes occur in the transition period from \Sexpr{yrs[2]}
to \Sexpr{yrs[3]}, encompassing both the completion of transformation from
planned to free market economies and the intended entry into the EMU.
During this time, the countries were successful in
decreasing all three moments of their inflation distributions while
simultaneously decreasing the heterogeneity (as captured by the corresponding
interquartile ranges) across countries. Before the transition, all three
moments were extremely large compared to the other two groups, while afterwards
all moments are much closer to those of the other groups. In Jan~\Sexpr{yrs[3]},
mean inflation is in between the levels for the core euro and PIIGS countries,
variance is still slightly higher, and skewness is not too far away from symmetry.

In context of the convergence issue, we find that up to \Sexpr{yrs[3]}
the inflation dynamics of all three groups of countries become much
more similar, both within and across groups. The interquartile ranges
of the means decline and become tighter for all groups and there appears to
be convergence towards some common mean inflation rate in the
countries of our sample. Variances also move more closely together, corresponding
to an increase for the core euro and PIIGS countries but a clear decrease
for the others. Right-skewness (and thus large outlying inflation rates)
has been reduced for all three groups of countries but to a different extent.
Thus, while ending in similar inflation dynamics, the timing and patterns of
changes in this convergence process are rather different across
the three groups, depending on the different starting situations of the
countries.

@@

The results from the previous sections convey that the financial
crisis of 2008/2009 did not have a significant effect on the inflation dynamics
in most of the European countries (except Ireland and Spain).
This may be somewhat surprising and is not entirely correct. In fact,
there is a rather clear effect, but as it is very short-lived and not
persistent, it is not picked up by the structural break techniques.
